---
title: "ASV data"
format: html
editor: visual
number-sections: true
---

Below we will use the provided ASV data to combine two data frames and perform some typical calculations.

## Compile ASV and metadata

```{r}
#| echo: false
asv_table <- read.delim(file = "r-intro-docs/asv-table.txt")
metadata <- read.delim(file = "r-intro-docs/metadata.txt")
```


```{r}
library(tidyverse)
# head(metadata)
asv_long <- asv_table %>%
  pivot_longer(starts_with("Gorda"), names_to = "SAMPLEID") %>% 
  left_join(metadata %>% select(SAMPLEID = SAMPLE, everything()))

# View(asv_long)
```

## Get temperature data from data frame

Isolate samples that have a temperature greater than 15 C.

```{r}
asv_long %>% 
  filter(temp > 15)

# Make a list of vent sites from Gorda Ridge that had a temperature greater than 15 C.
```

What are the average temperatures at the vent, plume, versus background samples? And how many of each sample are there?

```{r}
#| eval: false

asv_long %>% 
```


```{r}
#| eval: false
#| echo: false
#| message: false

asv_long %>% 
  select(SAMPLETYPE, temp) %>% 
  distinct() %>% 
  group_by(SAMPLETYPE) %>% 
    summarise(AVG_TEMP = mean(temp),
              COUNT = n())
```


## Parse taxon names

```{r}
#| eval: false

asv_long %>% 
```



## Calculate relative abundance

Create a table with sequence counts and ASV counts by supergroup.

```{r}
#| eval: false

asv_long %>% 
```


### Make a bar plot
```{r}
#| eval: false

asv_long %>% 
```

