---
title: "Basic plots"
format: html
editor: visual
number-sections: true
---

Also see: https://github.com/biovcnet/topic-R/blob/master/Lesson-6/lesson-06-bvcn-full.R

## Bar
```{r}
#| message: false
library(tidyverse)
head(starwars)
```


### Star Wars visuals

What species are present on each planet?

Create a bar plot where the x axis equals homeworlds and the y axis show the number of different species on each planet.
```{r}
#| eval: false
#| echo: false

starwars %>% 
  # filter(sex == "female") %>%
  ggplot(aes(x = homeworld, fill = species)) +
    geom_bar(stat = "count", position = "stack")
```


## Point & line

For all humans in Star Wars, what is their relationship between height and weight? and is it impacted by their planet of origin?

```{r}
#| eval: false
starwars %>%
```


```{r}
#| eval: false
#| echo: false
head(starwars)
starwars %>% 
  filter(species == "Human") %>%
  ggplot(aes(x = mass, y = height, fill = homeworld)) +
    geom_point(shape = 21, color = "black", size = 4, aes(fill = homeworld))
```
```{r}
#| eval: false
#| echo: false
head(starwars)
starwars %>% 
  filter(species == "Human") %>%
  ggplot(aes(x = mass, y = height, fill = homeworld)) +
    geom_point(shape = 21, color = "black", size = 4, aes(fill = homeworld)) +
  theme_linedraw() +
  labs(title = "mass & height of humans in star wars")
```

## Boxplot

# First what is the data I'm looking at on Tatooine
```{r}
# View(starwars)

tatooine <- starwars %>% 
  filter(homeworld == "Tatooine") %>% # select only those from Tatooine
  data.frame
```


# How many species on Tatooine?

```{r}
unique(starwars$species); length(unique(starwars$species))
unique(tatooine$species); length(unique(tatooine$species))
```

# Bar chart example
## First, example of a bar chart **but something isn't right**

```{r}
ggplot(tatooine, aes(x = species, y = height)) +
  geom_bar(stat = "identity")
```

# What's wrong with this?

```{r}
starwars %>% 
  filter(homeworld == "Tatooine") %>%
  group_by(species, homeworld) %>%
  summarise(MEAN_height = mean(height), MEDIAN_height = median(height),
            MAX_height = max(height), MIN_height = min(height),
            MEAN_mass = mean(mass), MEDIAN_mass = median(mass))

hist((filter(tatooine, species == "Human"))$height)  

hist((filter(tatooine, species == "Droid"))$mass) 
```

A better way to show this data is via box plot!
Let's address that question now, but with a better graphical representation
```{r}
ggplot(tatooine, aes(x = species, y = height)) +
  geom_boxplot()
```


### Boxplots in ggplot:
* median at middle
* upper/lower hinges = 1st and 3rd quartiles (25th and 75th percentiles)
* whiskers = largest/lowest value, but maxes at 1.5 * inter-quartile range (distance from upper/lower hinges)
* Outliers are shown as points
* NOTE: varies from base R 'boxplot()'

```{r}
ggplot(tatooine, aes(x = species, y = height)) +
  geom_boxplot() +
  geom_point()
```


