---
title: "Data import"
editor: visual
number-sections: true
---

## Import data from `r-intro-docs`.

In the files section of your RStudio, click on the `r-intro-docs` directory. You should see a list of files.

Lets save the greetings R object as a csv file in the `r-intro-docs` directory.

```{r}
# First re-create if needed
greetings <- c("Hey", "Hi", "Howdy", "Hello", "Morning")
n <- c(99, 15, 324, 54, 23)
df <- data.frame(greetings, n)
addition <- c("Afternoon", 18)

greetings <- rbind(df, addition)
```

*Discuss tab completion!*

```{r}
write.csv(greetings, "r-intro-docs/greetings.csv")
```

Open up `greetings.csv` locally. What does it look like in Excel?

```{r}
#| eval: false
?write.csv()
?write.table()
```

### Provided datasets

Import pizza using `read.csv()`.

```{r}
#| eval: false
#| echo: false
pizza <- data.frame(type = c("Meat", "BBQ Chicken", "Hawaiian", "Veggie", "Cheese", "Pepperoni", "Margherita", "Three cheese"),
                    num_toppings = c("3", "2", "3", "5", "0", "1", "2", "3"),
                    red_sauce = c(1, 1, 1, 1, 1, 1, 1, 0), num_order = c(2, 0, 0, 3, 3, 2, 4, 0), cost = c(7.5, 9, 7.5, 7.5, 6.5, 7.5, 7.5, 8.5))

write.csv(pizza, file = "r-intro-docs/pizza.csv")
```

```{r}
pizza <- read.csv("r-intro-docs/pizza.csv") 
```

### Star Wars data

Within the Tidyverse there is a Star Wars dataset we will use to explore data manipulation and wrangling in R.

Ensure Tidyverse is installed.

```{r}
#| eval: false
library(tidyverse)

data("starwars")

View(starwars)
```

### Activity

> Choose Star Wars or Pizza and describe the data types.

```{r}
#| echo: false
#| eval: false
str(pizza)
str(starwars)
```

### Import larger data frames

Data is from an ASV table from [Hu et al. 2022](https://shu251.github.io/microeuk-amplicon-survey/).

```{r}
#| eval: false
#| echo: false
#| message: false

load(file = "../../Projects/microeuks_deepbiosphere_datamine/microeuk-amplicon-survey/data-input/asv-tables-processed-27012022.RData", verbose = TRUE)

# head(asv_insitu)
asv_table <- asv_insitu_qc %>% 
  select(FeatureID, SAMPLE, value, Taxon) %>% 
  filter(grepl("Gorda", SAMPLE)) %>% 
  pivot_wider(names_from = "SAMPLE", values_from = "value", values_fill = 0)
# head(asv_table)

write_delim(asv_table, file = "r-intro-docs/asv-table.txt", delim = "\t")

metadata <- read_delim(file = "../../Projects/microeuks_deepbiosphere_datamine/microeuk-amplicon-survey/data-input/samplelist-metadata.txt")
metadata <- metadata %>% 
  select(SAMPLE, VENT, SITE, Sample_or_Control, SAMPLETYPE, DEPTH, temp)

write_delim(metadata, file = "r-intro-docs/metadata.txt", delim = "\t")


# run for class
asv_table <- read.delim(file = "r-intro-docs/asv-table.txt")
metadata <- read.delim(file = "r-intro-docs/metadata.txt")
```

Import the ASV table and associated metadata table that are in the `r-intro-docs` directory. Set them equal to **asv_table** and **metadata**, respectively.

```{r}
#| eval: false
asv_table <- 

metadata <-
```

Use the command `summary()` to learn more about the data frames we've imported above.

```{r}
#| eval: false
summary(asv_table)
head(asv_table)
head(metadata)
```
We will come back to this larger dataset after we explore some tidyverse functions.