---
title: "Star Wars data"
format: html
editor: visual
number-sections: true
---

## Explore the Star Wars data in tidyverse

```{r}
#| message: false
library(tidyverse)
head(starwars)
```

## Long vs. wide format

```{r}
#| eval: false
tatooine <- starwars %>% 
  filter(homeworld == "Tatooine") # select only those from Tatooine
```

*tibble*: is a dataframe specific version for tidyverse. It is both an object and a tibble. So it talks well with all of the related packages. Columns in a tibble are character vectors (instead of factors). This is technically easier to deal with.

#### Activity

> How many species on Tatooine?

```{r}
#| eval: false
unique(starwars$species)
length(unique(starwars$species))
```

Use your tidyverse cheatsheets and google!

#### Activity

> Create a table of droids that are equal to or greater than 96 inches in height.

```{r}
#| eval: false
#| echo: false
starwars %>% 
  filter(species == "Droid") %>% 
  filter(height >= 96)
```

## Summarizing data

What are the average masses of humans versus ewoks?

Use `group_by()` and `summarize()`.

```{r}
#| eval: false
starwars %>% 
  filter(species == "Human" | species == "Ewok") %>% 
  group_by(species) %>% 
  summarise(MEAN_mass = mean(mass))
# Modify the above to remove NAs and get a values for Humans.
```

What about the average mass of humans and Ewoks on different planets?

```{r}
#| eval: false
starwars %>% 
  filter(species == "Human" | species == "Ewok") %>% 
  group_by(species, homeworld) %>% 
  summarise(MEAN_mass = mean(mass))
```

`summarize()` creates a smaller output table, or a summary. This is dictated by `group_by()`. `Mutate()` adds a column to the data.

#### Activity

> Add a column to star wars that classifies tall versus short, based on the height of each species.

> Add a column that defines if the species is Human versus non-human.

> What are the mean masses of humans versus non-humans? And how many examples are included in this table.

> What are the max and min heights of humans and non-humans by planet?


## Star wars data wrangling

Isolate non-droids on Alderaan, Naboo, Endor, Kamino, and Coruscant.
Based on their reported sex, is there a relationship between their heigh, planet, and species?
```{r}

head(starwars)
unique(starwars$homeworld)

homes <- c("Alderaan", "Naboo", "Endor", "Kamino", "Coruscant")

starwars %>% 
  filter(homeworld %in% homes) %>% 
  filter(species != "Droid") %>% 
  ggplot(aes(x = species, y = height, fill = sex)) +
    # geom_bar(stat = "identity", aes(fill = sex)) +
    geom_bar(stat = "identity", position = "dodge", aes(fill = sex)) +
    facet_grid(homeworld ~ .)
```

